Adversarial Training of Neural Networks\\ against Systematic Uncertainty

### 2016-08-22 paper.tex
save

\glnote{Distinction between statistical and systematic uncertainty.}
\glnote{Define nuisance parameters.}
\glnote{We want to build an accurate classifier whose output remains invariant with
respect to systematic uncertainties.}
% https://www.slac.stanford.edu/econf/C030908/papers/TUAT004.pdf
\glnote{Motivate the criterion: allow to derive guarantees no matter $\lambda$,
while $f$ may otherwise be good or very bad depending on the nuisance if trained
on the mixture. E.g. to form hypothesis to be later confirmed by data.}

@@

Let assume a probability space $(\Omega, {\cal F}, P)$, where $\Omega$ is a
sample space, ${\cal F}$ is a set of events and $P$ is a probability measure.
Let consider the multivariate random variables $X_\lambda: \Omega \mapsto
\mathbb{R}^p$ and $Y: \Omega \mapsto {\cal Y}$, where $X_\lambda$ depends on a
nuisance parameter $\lambda \in \mathbb{R}$ whose values define a continuously
parameterized family of its systematic uncertainties. That is, $X_\lambda$ and
$Y$ induce together a joint probability distribution $p(X,Y|\lambda)$, where the
conditional on $\lambda$ denotes $X_\lambda$. For training, let further assume a
finite set $\{ x_i, y_i, \lambda_i \}_{i=1}^N$ of realizations
$X_{\lambda_i}(\omega_i), Y(\omega_i)$, for $\omega_i \in \Omega$ and known
values $\lambda_i$ of the nuisance parameter. Our goal is to learn a function
$f(\cdot;\theta_f) : \mathbb{R}^p \mapsto {\cal Y}$ of parameters $\theta_f$
(e.g., a neural network-based classifier if ${\cal Y}$ is a finite set of
classes) and minimizing  a loss ${\cal L}_f(\theta_f)$. In addition, we require
that $f(X_\lambda ; \theta_f)$ should be robust to the value of the nuisance parameter $\lambda$ --
which remains unknown at test time. More specifically, we aim at building $f$
such that in the ideal case
\begin{equation}\label{eqn:criterion-true}
f(X_{\lambda_i}(\omega) ; \theta_f) = f(X_{\lambda_j}(\omega) ; \theta_f)
\end{equation} for any
sample $\omega \in \Omega$ and any $\lambda_i, \lambda_j$ pair of values of the
nuisance parameter.

Since we do not have training tuples $(X_{\lambda_i}(\omega),
X_{\lambda_j}(\omega))$ (for the same unknown $\omega$), we propose instead to
solve the closely related problem of finding a predictive function $f$ such that
\begin{equation}\label{eqn:criterion}
    P(\{ \omega | f(X_{\lambda_i}(\omega) ; \theta_f) = y \} ) = P( \{ \omega' | f(X_{\lambda_j}(\omega') ; \theta_f) = y\}) \text{ for all $y \in {\cal Y}$}.
\end{equation}
In words, we are looking for a predictive function $f$ such that  the
distribution of $f(X_\lambda; \theta_f)$ is invariant with respect to the nuisance
parameter $\lambda$. Note that a function $f$ for which Eqn.~\ref{eqn:criterion-true} is
true necessarily satisfies Eqn.~\ref{eqn:criterion}. The converse is however in
general not true, since the sets of samples $\{ \omega | f(X_{\lambda_i}(\omega); \theta_f) = y \}$
and $\{ \omega' | f(X_{\lambda_j}(\omega'); \theta_f) = y \}$ do not need to be the same
for the equality to hold.

\glnote{An alternative objective, but which leads to something
significantly different, is to aim at getting the best model adapted
to the unknown $\lambda$, possibly using predictive information from
regions of the input space affected by the nuisance parameter.}

@@

In this work, we repurpose adversarial training as a means to regularize the
predictive model $f$ in order to satisfy Eqn.~\ref{eqn:criterion}. In
particular, we pit $f$ against an adversary regression model $r(\cdot ;
\theta_r) : \mathbb{R} \mapsto \mathbb{R}$ of parameters $\theta_r$ and
associated loss ${\cal L}_r(\theta_r)$. This regressor takes as input realizations of $f(X_\lambda; \theta_f)$
and produces as output predictions $\smash{\hat\lambda} = r(f(X_\lambda; \theta_f))$
of the nuisance parameter. If $p(f(X_\lambda; \theta_f))$ varies with $\lambda$,
then the corresponding correlation can be captured by $r$. By contrast, if
$p(f(X_\lambda; \theta_f))$ is invariant with $\lambda$ as we require, then $r$
should perform poorly. Training $f$ such that it additionally minimizes the performance of $r$
therefore acts as a regularization towards Eqn.~\ref{eqn:criterion}.

>>>>

\glnote{Distinction between statistical and systematic uncertainty.}
\glnote{Define nuisance parameters.}
\glnote{We want to build an accurate classifier whose output remains invariant with
respect to systematic uncertainties.}
% https://www.slac.stanford.edu/econf/C030908/papers/TUAT004.pdf
\glnote{Motivate the criterion (which may not be obvious for the ML crowd). See pivotal quantity motivation.}

@@

Let assume a probability space $(\Omega, {\cal F}, P)$, where $\Omega$ is a
sample space, ${\cal F}$ is a set of events and $P$ is a probability measure.
Let consider the multivariate random variables $X_\lambda: \Omega \mapsto
\mathbb{R}^p$ and $Y: \Omega \mapsto {\cal Y}$, where $X_\lambda$ depends on a
nuisance parameter $\lambda \in \Lambda \subseteq \mathbb{R}$ whose values define a
parameterized family of its systematic uncertainties. That is, $X_\lambda$ and
$Y$ induce together a joint probability distribution $p(X,Y|\lambda)$, where the
conditional on $\lambda$ denotes $X_\lambda$. For training, let further assume a
finite set $\{ x_i, y_i, \lambda_i \}_{i=1}^N$ of realizations
$X_{\lambda_i}(\omega_i), Y(\omega_i)$, for $\omega_i \in \Omega$ and known
values $\lambda_i$ of the nuisance parameter. Our goal is to learn a function
$f(\cdot;\theta_f) : \mathbb{R}^p \mapsto {\cal Y}$ of parameters $\theta_f$
(e.g., a neural network-based classifier if ${\cal Y}$ is a finite set of
classes) and minimizing  a loss ${\cal L}_f(\theta_f)$. In addition, we require
that $f(X_\lambda ; \theta_f)$ should be robust to the value of the nuisance parameter $\lambda$ --
which remains unknown at test time. More specifically, we aim at building $f$
such that in the ideal case
\begin{equation}\label{eqn:criterion-true}
f(X_{\lambda_i}(\omega) ; \theta_f) = f(X_{\lambda_j}(\omega) ; \theta_f)
\end{equation} for any
sample $\omega \in \Omega$ and any $\lambda_i, \lambda_j$ pair of values of the
nuisance parameter.

Since we do not have training tuples $(X_{\lambda_i}(\omega),
X_{\lambda_j}(\omega))$ (for the same unknown $\omega$), we propose instead to
solve the closely related problem of finding a predictive function $f$ such that
\begin{equation}\label{eqn:criterion}
    P(\{ \omega | f(X_{\lambda_i}(\omega) ; \theta_f) = y \} ) = P( \{ \omega' | f(X_{\lambda_j}(\omega') ; \theta_f) = y\}) \text{ for all $y \in {\cal Y}$}.
\end{equation}
In words, we are looking for a predictive function $f$ which is a pivotal quantity with respect to the nuisance parameter.
That is, such that  the
distribution of $f(X_\lambda; \theta_f)$ is invariant with respect to $\lambda$. Note that a function $f$ for which Eqn.~\ref{eqn:criterion-true} is
true necessarily satisfies Eqn.~\ref{eqn:criterion}. The converse is however in
general not true, since the sets of samples $\{ \omega | f(X_{\lambda_i}(\omega); \theta_f) = y \}$
and $\{ \omega' | f(X_{\lambda_j}(\omega'); \theta_f) = y \}$ do not need to be the same
for the equality to hold.

@@

In this work, we repurpose adversarial training as a means to regularize the
predictive model $f$ in order to satisfy Eqn.~\ref{eqn:criterion}. In
particular, we pit $f$ against an adversary model $r(\cdot ;
\theta_r) : \mathbb{R} \mapsto \Lambda$ of parameters $\theta_r$ and
associated loss ${\cal L}_r(\theta_r)$. This model takes as input realizations of $f(X_\lambda; \theta_f)$
and produces as output predictions $\smash{\hat\lambda} = r(f(X_\lambda; \theta_f))$
of the nuisance parameter. If $p(f(X_\lambda; \theta_f))$ varies with $\lambda$,
then the corresponding correlation can be captured by $r$. By contrast, if
$p(f(X_\lambda; \theta_f))$ is invariant with $\lambda$ as we require, then $r$
should perform poorly. Training $f$ such that it additionally minimizes the performance of $r$
therefore acts as a regularization towards Eqn.~\ref{eqn:criterion}.
